---
title: "Question 3 EDA"
author: "Andrew Dawson"
output: pdf_document
bibliography: ref.bib
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F)

#import Libraries and Data
require(tidyverse)
require(gghighlight)
require(knitr)
load("Data/cleanedData.RData")
labs <- setdiff(knitr::all_labels(), "setup")
```


# Comparing Disparities in Educational Access

## Wage 

```{r lookWage}
## lets make a bar plot
#make the lines and means
DataComp <- Data %>% filter(Indicator == "4.1.2", Units == "PERCENT")
linesWage <- DataComp %>% 
  group_by(`Education level`, Quantile) %>% 
  summarise(quantileMean = mean(Value))
linesWage2 <- DataComp %>%
  group_by(`Education level`) %>%
  summarise(eduMean = mean(Value))
pointsWage <- DataComp %>% 
  group_by(`Education level`, Quantile, region) %>% 
  summarise(regionalMean = mean(Value)) %>% 
  left_join(linesWage, by = c("Education level" = "Education level", "Quantile" = "Quantile")) %>%
  left_join(linesWage2, by = c("Education level" = "Education level"))

#plot the data
pointsWage %>% ggplot(aes(region, regionalMean, fill = region))+
  geom_col()+
  facet_grid(`Education level` ~ Quantile, labeller = labeller("Education level"=c("LOWSEC"="Lower Secondary", "PRIMAR" = "Primary")))+
  gghighlight(regionalMean < quantileMean, calculate_per_facet = T)+
  #geom_hline(aes(yintercept = eduMean), lty = 3, colour="blue")+
  geom_hline(aes(yintercept = quantileMean), lty=2, colour="red")+
  theme(legend.position = "none")+
  tiltXText+
  labs(x = "Region", y="Rate of Completion", title = "Comparing Wealth with Completion Rate")
```

<!-- results of the plot -->

Within this plot, the general regions are compared with wealth quantiles, i.e. Q1 indicates the poorest of the population, and Q5 the wealthiest. 

<!-- Discussion/Critical Evaluation -->

The mean completion rate per region seems to almost have more wealthy people lack behind. I don't think this is a fair comparision so adjustment of the graph is needed.

Take this plot with a grain of salt, for example look at Europe for Q5. IQR is the entire range, and the median is 100 yet the mean is less than the grandmean. 

```{r lookWealthpt2}
#plot the data
pointsWage %>% ggplot(aes(region, regionalMean, fill = region))+
  geom_col()+
  facet_grid(`Education level` ~ Quantile, labeller = labeller("Education level"=c("LOWSEC"="Lower Secondary", "PRIMAR" = "Primary")))+
  gghighlight(regionalMean < eduMean, calculate_per_facet = T)+
  geom_hline(aes(yintercept = eduMean), lty = 3, colour="blue")+
  #geom_hline(aes(yintercept = quantileMean), lty=2, colour="red")+
  theme(legend.position = "none")+
  tiltXText+
  labs(x = "Region", y="Percentage of Children Achiving Min Pro", title = "Comparing Wealth with Completion Rate")
```

This plot better highlights the wealth disparity, as it highlights the poorer countries to have more issues in educational completion rate, which is known to be historically proven <!-- find another source -->

## Recommendation

Sexes may not exactly demonstrate a need to differentiate, however there may be bias within the data. 

Wealth shows that less fortunate families may need further support. 


# References

<div id="refs"></div>

# Appendix

```{r, ref.label=labs, echo=T, eval=F}
```
