# Disparities Within Educational Access

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = F, message = F, warning = F, fig.width=9.5, fig.height=6)

# Install necessary libraries
require(tidyverse)
require(gghighlight)
require(knitr)
require(kableExtra)

```

## Wage 

```{r lookWage, fig.cap="Sourced From SDG4 Global Database (Indicator 4.1,2)"}
##make the lines and means
#set up the data
DataComp <- Data %>% filter(Indicator == "4.1.2", Units == "PERCENT", !is.na(Continent))

#set up the 
linesWage <- DataComp %>% 
  group_by(`Education level`, Quantile) %>% 
  summarise(quantileMean = mean(Value))
linesWage2 <- DataComp %>%
  group_by(`Education level`) %>%
  summarise(eduMean = mean(Value))

#join all together
pointsWage <- DataComp %>% 
  group_by(`Education level`, Quantile, Continent) %>% 
  summarise(regionalMean = mean(Value)) %>% 
  left_join(linesWage, by = c("Education level" = "Education level", "Quantile" = "Quantile")) %>%
  left_join(linesWage2, by = c("Education level" = "Education level"))
#reorder data
pointsWage$`Education level` <- factor(pointsWage$`Education level`, levels = c("PRIMAR", "LOWSEC", "UPPSEC")) 

#plot the data
pointsWage %>% ggplot(aes(Continent, regionalMean, fill = Continent))+
  geom_col()+
  facet_grid(`Education level` ~ Quantile, labeller = labeller("Education level"=c("PRIMAR" = "Primary", "LOWSEC"="L. Second", "UPPSEC" = "U. Second"), "Quantile" = c("_T" = "Total")))+
  gghighlight(regionalMean < eduMean, calculate_per_facet = T)+
  geom_hline(aes(yintercept = eduMean), lty = 3, colour="blue")+
  theme(legend.position = "none")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  labs(x = "Region", 
       y="Mean Rate of Completion", 
       title = "Comparing Wealth with Completion Rate", 
       subtitle = "Coloured bars indicate below the mean")
```

<!-- results of the plot -->

In Figure 3.1, blue dotted lines mark the global means for Primary (80), Lower Secondary (66), and Upper Secondary (47) education. Coloured bars indicate continents performing below these means.

Africa consistently falls below the mean across all wealth quantiles and education levels. Notably, in Primary education, the wealthiest and poorest quantiles show similar regional trends -- this is the only instance where this occurs. For Secondary levels, only Africa remains below the mean in the wealthiest quantiles (Q4–Q5), whereas nearly all regions (excluding Europe in Lower Secondary) fall below it in the poorest. Oceania generally trails the mean in Q1–Q3, though for Primary schooling, it is below average in all quantiles.

<!-- Discussion/Critical Evaluation -->

When looking across the different wealth quantiles, it is important to account for multiple things; for instance, when looking at the different education levels they are not independent of another. Education levels are sequential; lower completion rates at higher levels reflect the requirement to pass previous stages

Another factor is informal education, defined by @OKState as "learning about cultural values, norms, and expected behaviours by participating in a society", this affects the quantiles as when a child engages within their respective culture they may find a shift away from education towards something more culturally important -- e.g. working within the family farm, supporting their parents or grandparents, etc. 

Finally, Figure 3.1 is not a perfect representation of the wealth, as it only captures the mean rate of completion. One cannot see the spread of information or outliers which may affect the mean, thus changing the information of the plot; however, this may be accounted for using weighted means instead. 

\newpage

## Gender Parity Analysis

```{r genderPlot, fig.height=6, fig.cap="Sourced From SDG4 Global Database (Indicator 4.5.1)"} 

# Filter data to only handle data of indicator 4.5.1 for parity analysis
q3_4.5 <- filter(Data, Indicator == "4.5.1") %>%
          # Standardizing all values into ratios
          mutate(Value = as.numeric(Value),
          Value_ratio = case_when(Units == "PERCENT" ~ Value / 100,
                                  Units == "Ratio" ~ Value))

# For analysis purpose, reorder education levels
q3_4.5$"Education level" <- factor(q3_4.5$"Education level", 
                                    levels = c("PREPRI", "PRIMAR", "LOWSEC", "UPPSEC", "SECOND"))

# Location Parity Analysis

# Filter out data needed to calculate gender parity index for completion rate
genderData <- subset(q3_4.5, 
                     SeriesDescription == "Adjusted gender parity index for completion rate, by location, wealth quintile and education level",
                     select = c(`Education level`, Value_ratio, SubRegion, Continent))
  
# Produce plot of gender parity according to education levels over the sub-regions
ganderPlot <- ggplot(data = genderData, aes(x = factor(SubRegion), y = Value_ratio)) +
              geom_boxplot(aes(fill = Continent)) +
              facet_grid(`Education level` ~ ., 
                          labeller = labeller("Education level" = c("PRIMAR" = "Primary",
                                                                    "LOWSEC" = "Lower Secondary", 
                                                                    "UPPSEC" = "Upper Secondary"))) +

              # Add threshold line for parity (=1 where equality happens)
              geom_hline(yintercept = 1, 
                         colour = "red",
                         linetype = "dashed") +
  
              # Point out median parity by education level and year
              stat_summary(fun = median, 
                           shape = 19, 
                           size = 0.1, 
                           color = "red") +
  
              # Improve plot labels and theme
              labs(title = "Gender parity index for completion rate",
                   x = "Sub-Region",
                   y = "Parity Female/Male Index",
                   subtitle = "Adjusted gender parity index for completion rate") +
              theme(plot.title = element_text(hjust = 0.5, size = 17),
                    axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1)) +
              scale_fill_brewer(palette = "Pastel1")
print(ganderPlot)

```

Figure 3.2 illustrates the Gender Parity Index (GPI), where a red dashed line at 1 represents equality and red points denote median completion rates.

In Oceania, Melanesia maintains parity in primary and lower secondary education but shifts to favour males in upper secondary. Micronesia consistently favours females, while Polynesia -- despite a wide range of values -- also shows a median preference for females across all levels. African sub-regions exhibit contrasting trends: Northern Africa increasingly favours females as education levels rise, whereas Sub-Saharan Africa sees a steady decline in parity at higher levels. This is explained further in section 1. 
 
\newpage

## Location Parity Analysis

```{r locationPlot} 
# Location Parity Analysis
 
# Filter out data needed to calculate location parity index for completion rate
locationData <- subset(q3_4.5, 
                       SeriesDescription == "Adjusted location parity index for completion rate, by sex, wealth quintile and education level",
                       select = c(SubRegion, `Education level`, Value_ratio, Continent, TimePeriod))
# Build a table of location parity according to education levels over the sub-regions
locationTable <- group_by(locationData, `Education level`, Continent) %>%
                  # Calculate measurements of data
                  summarise(Median = round(median(Value_ratio), 2),
                            Mean = round(mean(Value_ratio), 2),
                            Q1 = round(quantile(Value_ratio, 0.25), 2),
                            Q3 = round(quantile(Value_ratio, 0.75), 2),
                            .groups = "drop") %>%
                  # Improve table labels
                  mutate(`Education level` = recode(`Education level`,
                                                     "PRIMAR" = "Primary",
                                                     "LOWSEC" = "Lower Secondary", 
                                                     "UPPSEC" = "Upper Secondary"),
                         `IQR (Q1-Q3)` = paste0(Q1, " - ", Q3)) %>%
                  select(-Q1, -Q3)
 
# Display the table of location parity
gradient_colors <- colorRampPalette(c("darkgray", "white"))(100)
kable(locationTable, align = c("l","l","c","c","c"),
      caption = "Location parity index (Rural/Urban) for education completion rate") %>%
  column_spec(3, background = gradient_colors[locationTable$Median*99 + 1]) %>%
  column_spec(4, background = gradient_colors[locationTable$Mean*99 + 1]) %>%
  column_spec(5, extra_css = "text-align:center;") %>%
  collapse_rows(columns = 1, valign = "middle")  %>%
  add_header_above(c("GDB indincator 4.5.1" = 5), bold = FALSE)
```

Similar to Figure 3.2, Table 3.1 presents the Location Parity Index, where a value of 1 represents rural-urban equality. Darker cell shading indicates greater deviation from this threshold

Africa and Oceania demonstrate a consistent urban advantage across all education levels, with rural completion rates dropping further as education levels rise. In contrast, the Americas and Asia show a mixed distribution: roughly half the regions achieve parity -- with medians near 1 -- while the other half favour urban areas. Europe follows a similar split pattern; however, the group failing to reach parity is characterized by notably lower means and wider IQRs, indicating significant inequality within those specific regions.
 
## Recommendations 

Through the analysis of wealth, gender, and location disparities Africa consistently lags behind other regions. Consequently, looking towards including a thorough audit of educational resources and the implementation of specific equality strategies within Northern and Sub-Saharan Africa. Furthermore, rural areas require targeted training to strengthen learning environments. Disparities are also evident in Oceania. In Polynesia, mentorship and community awareness initiatives are recommended to address declines in female enrolment. Finally, broad investment in rural infrastructure -- specifically transportation and digital tools -- is essential to improve educational access.