---
title: "Question 1 EDA"
author: "Andrew Dawson"
output: pdf_document
bibliography: ref.bib
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

#import Libraries and Data
library(tidyverse)
load("cleanedData.RData")
labs <- setdiff(knitr::all_labels(), "setup")
```


## EDA and Analysis of 4.1.1

```{r lookGeneral}
#for future plots
tiltXText <- theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

#filter data
DataPro <- Data %>% filter(Indicator == "4.1.1", Units == "PERCENT", Sex == "BOTHSEX", `sub-region` != "NA")

#plot filtered data
DataPro %>% ggplot(aes(`sub-region`, Value))+
  geom_boxplot()+
  facet_wrap(~`Education level`)+
  gghighlight(quantile(Value,0.5)<75, calculate_per_facet = T)+
  tiltXText
```

Within the box plots, the highlighted box plots show within the differing sub-regions' median of population, which percentage of the population have not achieved minimum proficiency in reading and mathematics. This demonstrates the sub-regions capability of providing education access as the lower the percentage of the population with the minimum proficiency of the two vital skills, the lower the access is. 


Comparing this with the two sexes also shows different patterns. 

```{r lookSex}
#filter data
DataProSex <- Data %>% filter(Indicator == "4.1.1", Units == "PERCENT", Sex != "BOTHSEX", `sub-region` != "NA")

#plot filtered data
DataProSex %>% ggplot(aes(`sub-region`, Value))+
  geom_boxplot()+
  facet_grid(Sex~`Education level`)+
  gghighlight(quantile(Value,0.5)<75, calculate_per_facet = T)+
  tiltXText
```

Now, what is most concerning is when looking at Latin America and the Caribbean and Northern Africa. Both of these sub-regions in both plots demonstrate a general pattern that the percentage of the entire population -- or just about -- has not achieved 75% proficiency. 

```{r LA.NA.Table}

```



# References

<div id="refs"></div>

# Appendix

```{r, ref.label=labs, echo=T, eval=F}
```

