---
title: "Question 1 EDA"
author: "Andrew Dawson"
output: pdf_document
bibliography: ref.bib
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

#import Libraries and Data
require(tidyverse)
require(gghighlight)
require(knitr)
load("Data/cleanedData.RData")
labs <- setdiff(knitr::all_labels(), "setup")
```


## EDA and Analysis of 4.1.1

```{r lookGeneral}
#for future plots
tiltXText <- theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

#filter data
DataPro <- Data %>% filter(Indicator == "4.1.1", Units == "PERCENT", Sex == "BOTHSEX", `sub-region` != "NA")

#plot filtered data
DataPro %>% ggplot(aes(`sub-region`, Value))+
  geom_boxplot()+
  facet_wrap(~`Education level`)+
  gghighlight(quantile(Value,0.5)<75, calculate_per_facet = T)+
  tiltXText
```

Within the box plots, the highlighted box plots show within the differing sub-regions' median of population, which percentage of the population have not achieved minimum proficiency in reading and mathematics. This demonstrates the sub-regions capability of providing education access as the lower the percentage of the population with the minimum proficiency of the two vital skills, the lower the access is. 


```{r LA.NA.Table}
#table
tab <- DataPro %>% 
  arrange(Value) %>%
  select(GeoAreaName, `sub-region`, TimePeriod, Value, `Education level`, `Type of skill`) %>%
  slice_head(n=10)

kable(tab)
```

These countries display the worst percentages of their population, this aligns well with the plot.

# References

<div id="refs"></div>

# Appendix

```{r, ref.label=labs, echo=T, eval=F}
```

